msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-08 paper.tex
Updated discussion of recombination.

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
%\usepackage{showkeys}

@@

@@

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has profoundly changed
what is possible. Through the use of efficient data structures, it is
now possible to simulate millions of whole chromosomes, without resorting
to approximations. The ``succinct tree
sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} based formats. This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\subsection*{Recombination and gene conversion}
Recombination is implemented in \msprime\ via
Hudson's algorithm, which
works backwards in time tracking the
through the effects of common ancestor and recombination events
~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}. Common
ancestor events combine the ancestral material of two lineages, and may
 result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.

We also support variable recombination rates along a chromosome.
This allows us to model both recombination hotspots like
\texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
recombination maps like those supported \texttt{cosi2}
\citep{shlyakhter2014cosi2}
(What about ~\cite{wang2014new}?)
We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms.
% % Uncomment this sentence, when the corresponding issue in the msprime github repo is resolved.
% Furthermore, msprime is now able to simulate gene conversion events that are
% initiated proportional to the local recombination rate specified by the recombination map.

@@

\subsection*{SMC approximations}

\textbf{TODO:} It's important that we acknowledge that (a) the coalescent
with recombination does become quite slow for large $N_e$; (b) our
implementation of the SMC is naive; and (c) for sufficiently large
$N_e$ programs like \texttt{scrm} will be faster.

The SMC~\citep{mcvean2005approximating} and SMC'~\citep{marjoram2006fast}
models are supported.
The performance of msprime has made the SMC
approximation largely unnecessary, at least when simulating human-like
recombination rates. The SMC is an important analytical approximation
forming the basis for several important inference methods
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust},
so simulations are still useful to examine  how good the approximation
is.

The SMC has been shown to be a good approximation to the
coalescent with recombination for samples of less than
four~\citep{hobolth2014markovian,wilton2015smc}
and in simulations, produces patterns of LD indistinguishable from the
full coalescent~\citep{mcvean2005approximating,marjoram2006fast}. However,
LD is a crude approximation of the correlation structure of trees
and it is not clear how good the SMC is for larger sample sizes.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
%\usepackage{showkeys}

@@

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}

@@

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations.
Through the use of efficient data structures, it is
now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} or
VCF~\citep{danecek2011variant} based formats formats.
This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

\subsection*{Recombination}

Recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots like \texttt{msHOT}~\citep{hellenthal2007mshot}
and empirical recombination maps similar to other
simulators~\citep[e.g.][]{shlyakhter2014cosi2}.

The Sequentially Markovian Coalescent (SMC) is an approximation of the
coalescent with recombination
model~\citep{mcvean2005approximating,marjoram2006fast},
and was originally motivated to a significant degree by the need to
simulate longer genomes than was possible using tools like \ms.
The SMC is a good approximation to the
coalescent with recombination when we have four or fewer sampled
genomes~\citep{hobolth2014markovian,wilton2015smc}, but the
effects of the approximation are less well understood for larger
sample sizes and several approaches have been proposed to
allow simulations more closely approximate the coalescent
with recombination~\citep{chen2009fast,wang2014new,staab2014scrm}.
The SMC is an important analytical approximation
and has allowed the development of numerous inference
methods~\citep{
li2011inference,
harris2013inferring,
sheehan2013estimating,
schiffels2014inferring,
carmi2014renewal,
rasmussen2014genome,
zheng2014bayesian,
terhorst2017robust}.

\begin{figure}
\caption{\label{fig-recombination-perf}Approximate run time for \msprime\
simulations as a function of scaled recombination rate.}
\end{figure}

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho$, and therefore for sufficiently large $N_e$, recombination rate
or long genome, efficient SMC implementations will be faster
that \msprime.

\jkcomment{We want to round out this discussion here by giving some
sort of rough argument for why Hudson's algorithm is quadratic, and discussing
to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
to simulate in how long. See issue 28.}

The SMC~\citep{mcvean2005approximating} and
SMC'~\citep{marjoram2006fast} models are available
in \msprime\ 1.0. However, they are currently implemented using a
naive rejection sampling approach, and are somewhat slower
to simulate than the exact coalescent with recombination. These
models are therefore only appropriate for studying the SMC approximations
themselves; an important avenue for future work is to implement
efficient SMC-like models, perhaps based on the approaches used
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.

\subsection*{Gene conversion}

We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms.

@@
